---
title: "SEM Analysis"
subtitle: "Alexander James Ryan"
output: pdf_document
toc: true
toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# List of Libraries
library(readr)
library(semPlot)
library(lavaan)

# Load dataset
data <- read_delim("~/Desktop/Structural-equation-modelling/data.csv", 
    "\t", escape_double = FALSE, trim_ws = TRUE)
# Create subset, removing unnessary variables
myvars <- c("race", "age", "engnat", "gender", "hand", "E1", "E2", "E3", "E4", "E5", "E6", "E7", "E8", "E9", "E10", "N1", "N2", "N3", "N4", "N5", "N6", "N7", "N8", "N9", "N10", "A1", "A2", "A3", "A4", "A5", "A6", "A7", "A8", "A9", "A10", "C1", "C2", "C3", "C4", "C5", "C6", "C7", "C8", "C9", "C10", "O1", "O2", "O3", "O4", "O5", "O6", "O7", "O8", "O9", "O10")
ds <- data[myvars]

# Record race, gender, engnat, hand as factors
ds <- within(ds, {
  race <- factor(race)
  gender <- factor(gender)
  engnat <- factor(engnat)
  hand <- factor(hand)
})

ds$gender[ds$gender==0] <- NA
ds$gender[ds$gender==3] <- NA
ds$gender[ds$gender==2] <- 0
```

## Introduction

A lot has been written about various psychology Factor Analysis models. In this example, a dataset will be used to show psycholoigcal traits in the latent variables. The list of questions asked have already been systemitcally tailored to derive a good estimate of an individual's latent personality structure. Ignoring the literature already devloped on the model, I hope to show how all these factors relate to one another, including the variables gender.

### Research Questions

What is the relationship between the Latent variables:
Extroversion, Openness, Neurotiscism, Agreeableness, Conscioustousness

Which questions from the questionaire have the biggest impact on their respective latent variable?

Do other questions have an impact on other latent variables?

Are there more than just 5 personality latent variables? -> how the hell should I test this?

### Dataset explanation

The dataset is an online questionaire conducted in 2012. A series of questions using Likert scaling were asked and each respondant had to rank their response from 1, they disagree, to 5, they agree (0 meant the question was missed). The questions were not asked in any particular order. Other questions that were asked, were the race of the person, a categorical value, the age of the participant, a binary variable if the partiicpant was fluent in English, the gender of the person, and finally their dominant writing hand. Acquiesence bias is managed by using questions with opposite meanings, for example, a person who answers agree to question "E5 I start conversations" is not going to also write agree to "E6 I have little to say".

# Dataset Exploration
```{r}
nrow(ds) # number of subjects = 19719
ncol(ds) # number of variables = 57
```


```{r}
cov(ds) # covariance matrix
```







## Model Specification

```{r}
model <- 'E =~ E1 + E2 + E3 + E4 + E5 + E6 + E7 + E8 + E9 + E10
N =~ N1 + N2 + N3 + N4 + N5 + N6 + N7 + N8 + N9 + N10
A =~ A1 + A2 + A3 + A4 + A5 + A6 + A7 + A8 + A9 + A10
C =~ C1 + C2 + C3 + C4 + C5 + C6 + C7 + C8 + C9 + C10
O =~ O1 + O2 + O3 + O4 + O5 + O6 + O7 + O8 + O9 + O10
E ~ gender
N ~ gender
A ~ gender
C ~ gender
O ~ gender' 
fit <- cfa(model, data=ds)

summary(fit, fit.measures = TRUE, standardized=TRUE)
parameterEstimates(fit)
resid(fit, type = "standardized")
fitted(fit)
fitMeasures(fit)
fitMeasures(fit, c("cfi", "rmsea", "srmr"))  #selection of fit measures
fitMeasures(fit, "bic") # gives only one
inspect(fit) # gives model matrices

semPaths(fit, "model", "stand", style ="LISREL", rotation=1, edge.color="black", edge.label.cex=1, mar=c(10,1,2,1))

```




# Model Assessment + Model Fit

## Comparative Fit Index (CFI) 

## Standardized Root Mean Square Residual (SRMR)

## Root Mean Square Error of Approximation (RMSEA)

## Apendix

### Questions asked

#### Personality Questions

E1	I am the life of the party.
E2	I don't talk a lot.
E3	I feel comfortable around people.
E4	I keep in the background.
E5	I start conversations.
E6	I have little to say.
E7	I talk to a lot of different people at parties.
E8	I don't like to draw attention to myself.
E9	I don't mind being the center of attention.
E10	I am quiet around strangers.
N1	I get stressed out easily.
N2	I am relaxed most of the time.
N3	I worry about things.
N4	I seldom feel blue.
N5	I am easily disturbed.
N6	I get upset easily.
N7	I change my mood a lot.
N8	I have frequent mood swings.
N9	I get irritated easily.
N10	I often feel blue.
A1	I feel little concern for others.
A2	I am interested in people.
A3	I insult people.
A4	I sympathize with others' feelings.
A5	I am not interested in other people's problems.
A6	I have a soft heart.
A7	I am not really interested in others.
A8	I take time out for others.
A9	I feel others' emotions.
A10	I make people feel at ease.
C1	I am always prepared.
C2	I leave my belongings around.
C3	I pay attention to details.
C4	I make a mess of things.
C5	I get chores done right away.
C6	I often forget to put things back in their proper place.
C7	I like order.
C8	I shirk my duties.
C9	I follow a schedule.
C10	I am exacting in my work.
O1	I have a rich vocabulary.
O2	I have difficulty understanding abstract ideas.
O3	I have a vivid imagination.
O4	I am not interested in abstract ideas.
O5	I have excellent ideas.
O6	I do not have a good imagination.
O7	I am quick to understand things.
O8	I use difficult words.
O9	I spend time reflecting on things.
O10	I am full of ideas.

#### Other Questions

race 

1 = Mixed Race, 2 = Arctic (Siberian, Eskimo), 3 = Caucasian (European), 4 = Caucasian (Indian), 5 = Caucasian (Middle East), 6 = Caucasian (North African, Other), 7 = Indigenous Australian, 8 = Native American, 9 = North East Asian (Mongol, Tibetan, Korean Japanese, etc), 10 = Pacific (Polynesian, Micronesian, etc), 11 = South East Asian (Chinese, Thai, Malay, Filipino, etc), 12 = West African, Bushmen, Ethiopian, 13 = Other (0=missed)



