---
title: "SEM Analysis"
subtitle: "Alexander James Ryan"
output: pdf_document
toc: true
toc_depth: 3
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# List of Libraries
library(readr)
library(semPlot)
library(lavaan)

# Load dataset
ds <- read_delim("~/Desktop/Structural-equation-modelling/data.csv", 
    ";", escape_double = FALSE, trim_ws = TRUE)

# Record race, gender, engnat, hand as factors
ds <- within(ds, {
  US <- factor(US)
  GB <- factor(GB)
  FR <- factor(FR)
  IT <- factor(IT)
  PK <- factor(PK)
  RO <- factor(RO)
  IN <- factor(IN)
  DK <- factor(DK)
  CA <- factor(CA)
  BE <- factor(BE)
  BR <- factor(BR)
  AU <- factor(AU)
  DE <- factor(DE)
  CN <- factor(CN)
  IE <- factor(IE)
  AT <- factor(AT)
  CZ <- factor(CZ)
  GR <- factor(GR)
  ID <- factor(ID)
  JP <- factor(JP)
  KR <- factor(KR)
  NL <- factor(NL)
  NO <- factor(NO)
  ES <- factor(ES)
  SE <- factor(SE)
  FI <- factor(FI)
  race <- factor(race)
  gender <- factor(gender)
  engnat <- factor(engnat)
  hand <- factor(hand)
})

ds$gender[ds$gender==0] <- NA # deletes this gender option
ds$gender[ds$gender==3] <- NA # deletes this gender option
ds$gender[ds$gender==2] <- 0 # changes women to value 0


# Lists for Models and Fits
models <- list()
fits <- list()


```

## Introduction

A lot has been written about various psychological personality models using Factor Analysis. In this example, a dataset will be used to show psycholoigcal traits in the latent variables. The list of questions in the survey have already been systemitcally tailored to derive a good estimate of an individual's latent personality structure. Ignoring the literature already devloped on the model, I hope to show how all these factors relate to one another, including the variables such as gender and country of origin.

## Research Questions

What is the relationship between the Latent variables:
Extroversion, Openness, Neurotiscism, Agreeableness, Conscioustousness

Which questions from the questionaire have the biggest impact on their respective latent variable?

Do other questions have an impact on other latent variables?

Are there more than just 5 personality latent variables? -> how the hell should I test this?

What effect does Gender have on the structure of the personality latent variables? The fundamental question, are there differences in personality structure between men and women.

What effect does age have on the distribution of an individuals personality?

How does one's country affect one's personality distribution? Are some countries higher or lower in personality factors than others?

## Dataset explanation

The dataset is an online questionaire conducted in 2012 by Goldberg et al., with 19719 participants. 50 different survey questions (called items) relating to personality using Likert scaling were asked and each respondant had to rank their response from 1, they disagree, to 5, they agree (0 meant the question was missed). The questions were not asked in any particular order. Other questions that were asked, were the race of the person, a categorical value, the age of the participant, a binary variable if the participant was fluent in English, the gender of the person, and finally their dominant writing hand. The IP address of the partipant was also recorded, so there is also a country of origin variable.

Each item is designed to reveal a different latent variable, the personality traits, Extroversion, Neuroticism, Agreeableness, Consciestounes, and Openness (shortened to ENACO).

Some critisms of the validity of the data include the presence of acquience bias, where survey participants have a tendency to agree with a disproportiate number of questions, irrespective of the content of the question. This is managed by using items with opposite meanings, for example, a person who answers agree to question "E5 I start conversations" is not going to also write agree to "E6 I have little to say".

Another critism is that the nature of the responses is self reported, that is, the participants are answering personality questions about themselves. This opens up the possibility of bias in the results. Participants may answer the questions in how they wish to be, rather how they actually feel.

# Dataset Exploration
```{r exploration, echo=FALSE, results=FALSE, warning=FALSE}
nrow(ds) # number of subjects = 19719
ncol(ds) # number of variables = 57
mean(ds$age) # mean age is ~26
median(ds$age) # median age is 22
```


## Model Specification

Restrictions are placed on E1, N1, A2, C1, O1, the items that positively correlated with each personality trait. This means that positive loadings relate to stronger values for that particular personality trait, whereas negative loadings imply there is a negative relation. For example, if Extraversion loads negatively (loading=-0.45) onto Neurotisim, this means that those higher in the trait Extraversion, will on average have lower levels of the trait Neurotisim.


### Five Factor Model

Each personality latent variable has its own items relating to that latent variable.

      E =~ E1 + E2 + E3 + E4 + E5 + E6 + E7 + E8 + E9 + E10  
      N =~ N1 + N2 + N3 + N4 + N5 + N6 + N7 + N8 + N9 + N10  
      A =~ A2 + A1 + A3 + A4 + A5 + A6 + A7 + A8 + A9 + A10  
      C =~ C1 + C2 + C3 + C4 + C5 + C6 + C7 + C8 + C9 + C10  
      O =~ O1 + O2 + O3 + O4 + O5 + O6 + O7 + O8 + O9 + O10  

Every item is loaded onto every latent variable

## Final Model

Each item loads onto its respective latent variable. The regressions include age and gender loading onto ENACO latent variables, with hand loading onto Agreeableness and Openness. Also included are residual correlations between items. These are items which are similarly worded. For example:  
N7	I change my mood a lot.  
N8	I have frequent mood swings.  



```{r models, echo=FALSE, results=FALSE, warning=FALSE}
# Model 1:  Five Factor Model 
# (A2 needs to go first, so agreeableness factor is positive for high agree)
  fits$m1 <- sem(models$m1, data=ds)
    summary(fits$m1, fit.measures = TRUE, standardized = TRUE)

models$m1 <- '
    # measurement model
      E =~ E1 + E2 + E3 + E4 + E5 + E6 + E7 + E8 + E9 + E10
      N =~ N1 + N2 + N3 + N4 + N5 + N6 + N7 + N8 + N9 + N10
      A =~ A2 + A1 + A3 + A4 + A5 + A6 + A7 + A8 + A9 + A10
      C =~ C1 + C2 + C3 + C4 + C5 + C6 + C7 + C8 + C9 + C10
      O =~ O1 + O2 + O3 + O4 + O5 + O6 + O7 + O8 + O9 + O10'

# Model 2:  Five Factor Model with age and gender and engnat and hand
  fits$m2 <- sem(models$m2, data=ds)
    summary(fits$m2, fit.measures = TRUE, standardized = TRUE)

models$m2 <- '
    # measurement model
      E =~ E1 + E2 + E3 + E4 + E5 + E6 + E7 + E8 + E9 + E10
      N =~ N1 + N2 + N3 + N4 + N5 + N6 + N7 + N8 + N9 + N10
      A =~ A2 + A1 + A3 + A4 + A5 + A6 + A7 + A8 + A9 + A10
      C =~ C1 + C2 + C3 + C4 + C5 + C6 + C7 + C8 + C9 + C10
      O =~ O1 + O2 + O3 + O4 + O5 + O6 + O7 + O8 + O9 + O10
    # regressions
      E ~ age + gender + hand
      N ~ age + gender + hand
      A ~ age + gender + hand
      C ~ age + gender + hand
      O ~ age + gender + hand
'

# Model 3:  Five Factor Model (with Country)
  fits$m3 <- sem(models$m3, data=ds)
    summary(fits$m3, fit.measures = TRUE, standardized = TRUE)

models$m3 <- '
    # measurement model
      E =~ E1 + E2 + E3 + E4 + E5 + E6 + E7 + E8 + E9 + E10
      N =~ N1 + N2 + N3 + N4 + N5 + N6 + N7 + N8 + N9 + N10
      A =~ A2 + A1 + A3 + A4 + A5 + A6 + A7 + A8 + A9 + A10
      C =~ C1 + C2 + C3 + C4 + C5 + C6 + C7 + C8 + C9 + C10
      O =~ O1 + O2 + O3 + O4 + O5 + O6 + O7 + O8 + O9 + O10
    # regressions
      E ~ US+GB+AU+IE+NL+NO++SE+IN+DK+CA+BE+BR
      N ~ US+GB+AU+DE+NL+NO+IN+DK+CA+BE+BR
      A ~ US+FR+AU+DE+IE+ID+IN+DK+CA+BE+BR
      C ~ US+GB+FR+IT+AU+DE+AT+ID+ES+SE+BR
      O ~ US+FR+IT+AU+DE+AT+ID+NL+NO+ES+RO+IN+DK+CA+BE+BR
' 

# Model 4:  Five Factor Model (with Race)
  fits$m4 <- sem(models$m4, data=ds)
    summary(fits$m4, fit.measures = TRUE, standardized = TRUE)
 
models$m4 <- '
    # measurement model
      E =~ E1 + E2 + E3 + E4 + E5 + E6 + E7 + E8 + E9 + E10
      N =~ N1 + N2 + N3 + N4 + N5 + N6 + N7 + N8 + N9 + N10
      A =~ A2 + A1 + A3 + A4 + A5 + A6 + A7 + A8 + A9 + A10
      C =~ C1 + C2 + C3 + C4 + C5 + C6 + C7 + C8 + C9 + C10
      O =~ O1 + O2 + O3 + O4 + O5 + O6 + O7 + O8 + O9 + O10
    # regressions
      E ~ r2 + r11
      N ~ r3 + r4 + r5 + r11 + r12
      A ~ r1 + r3 + r7 + r9 + r11
      C ~ r1 + r3 + r4 + r6 + r7 + r9
      O ~ r1 + r3 + r11 + r12
' 

# Model 5:  Five Factor Model (with correlated residuals)
  fits$m5 <- sem(models$m5, data=ds)
    summary(fits$m5, fit.measures = TRUE, standardized = TRUE)

models$m5 <- '
    # measurement model
      E =~ E1 + E2 + E3 + E4 + E5 + E6 + E7 + E8 + E9 + E10
      N =~ N1 + N2 + N3 + N4 + N5 + N6 + N7 + N8 + N9 + N10
      A =~ A2 + A1 + A3 + A4 + A5 + A6 + A7 + A8 + A9 + A10
      C =~ C1 + C2 + C3 + C4 + C5 + C6 + C7 + C8 + C9 + C10
      O =~ O1 + O2 + O3 + O4 + O5 + O6 + O7 + O8 + O9 + O10
    # residual correlations
      N7 ~~ N8
      O1 ~~ O8
      O2 ~~ O4
      O5 ~~ O10
      O3 ~~ O6
      E8 ~~ E9
      A2 ~~ A7
      A4 ~~ A9
      C2 ~~ C6
      E2 ~~ E6
      N1 ~~ N3
      N4 ~~ N10 
      A5 ~~ A7
      N9 ~~ A3
    # cross loadings
      N =~ E3
      O =~ C3
      C =~ A3
      E =~ A7
      E =~ A4
      E =~ A2 
      E =~ A10  
      N =~ C4
      O =~ C10
      O =~ E6
      N =~ O9
      N =~ A3
      N =~ A6
      A =~ E3
      N =~ O3
      E =~ N10
      A =~ N9'

# Model 6:  Final model
  fits$m6 <- sem(models$m6, data=ds)
    summary(fits$m6, fit.measures = TRUE, standardized = TRUE)

models$m6 <- '
    # measurement model
      E =~ E1 + E2 + E3 + E4 + E5 + E6 + E7 + E8 + E9 + E10
      N =~ N1 + N2 + N3 + N4 + N5 + N6 + N7 + N8 + N9 + N10
      A =~ A2 + A1 + A3 + A4 + A5 + A6 + A7 + A8 + A9 + A10
      C =~ C1 + C2 + C3 + C4 + C5 + C6 + C7 + C8 + C9 + C10
      O =~ O1 + O2 + O3 + O4 + O5 + O6 + O7 + O8 + O9 + O10
    # regressions
      E ~ age + gender
      N ~ age + gender
      A ~ age + gender + hand
      C ~ age + gender
      O ~ age + gender + hand
    # residual correlations
      N7 ~~ N8
      O1 ~~ O8
      O2 ~~ O4
      O5 ~~ O10
      O3 ~~ O6
      E8 ~~ E9
      A2 ~~ A7
      A4 ~~ A9
      C2 ~~ C6
      E2 ~~ E6
      N1 ~~ N3
      N4 ~~ N10 
      A5 ~~ A7
'

# Modification Indices
modificationindices(object = fits$m8, sort. = TRUE)


v <- list()

v$fitindicies <- c("npar", "chisq", "df", "pvalue", "cfi", "rmsea", 
                   "rmsea.ci.lower", "rmsea.ci.upper", "srmr")

# Core fit function
core_fitmeasure <- function(fit = fits$m1, fitindicies = v$fitindicies, digits = 3 ) {
  x <- fitMeasures(fit)
  round(x[fitindicies], digits)
}

# Create table with comparison of models
sapply(fits, function(X) core_fitmeasure(X))


# SEM Path
semPaths(fits$m4, "model", "stand", style ="LISREL", rotation=1, edge.color="black", edge.label.cex=1, mar=c(10,1,2,1))



```




# Model Assessment + Model Fit

## Comparative Fit Index (CFI) 

## Standardized Root Mean Square Residual (SRMR)

## Root Mean Square Error of Approximation (RMSEA)


# Model results

## Covariances between Personality Factors

Extraversion loads negatively onto Neurotisim (), and positively onto Agreeableness(), Conscioetousness () and Openness (). This implies those higher in trait Extraversion on average have lower levels of Neurotisim, and higher levels of Agreeableness, Conscientousness and Openness.

Neuroticsism loads negatively onto Agreeableness (), Conscientousness () and Openness (). Agreeableness loads positively onto Conscientousness () and Openness (). This implies that those higher in trait Neuroticsism, have on average lower levels of Agreeableness, Conscientousness and Openness.

Agreeableness loads positively onto Conscioustousness () and Openness (). This implies those with higher levels of the trait Agreeableness have on average higher levels of Consciousness and Openness.

Conscientousness loads positively onto Openness. This implies that those with higher levels of Consciousness have higher levels of Openness.

## Effect of Gender on personality factors

Gender, an exogenous variable, is loaded onto the latent factors ENACO as regression coefficients. Gender has a statistically significant loading onto each personality factor. The standardized latent variable for Extraversion loading is (-0.123), for Neurotisism (-0.373), for Agreeableness (-0.458), for Consciestiousness (-0.062), and Openness (0.257). These results imply that women on average have a lower level of Extraversion than men, a higher level of Neurotism than men, a higher level of Agreeableness than men, a marginally lower level of Conscioestiousness, and a lower level of Openness than men.

## Effect of Age on personality factors

Age was statistically significant for all the latent factors ENACO. Age appears to have a positive effect on Extraversion (), Agreeableness (), Conscioetouness () and Openness (), and a negative effect on Neuroticism (). This result implies that as people age, on average they will experience lower levels of negative emotion.

## Effect of dominate hand on personality factors

Dominate hand, an exogenous variable is loaded onto the latent factors ENACO as regression coefficients. For Extroversion, Neurotisicm, and Consciousness, the loading is not statistically significant. This implies that being left or right handed does not affect these personality characteristics. For Agreeableness the standardized latent variable is (-0.072) and for Openness it is (0.148). This implies that those who are left handed, or ambidexious, are on average lower in the trait Agreeableness, and also on average they are higher in the trait Openness.

## Effect of race upon personality factors

13 different categories for race were loaded onto ENACO as regression coefficients. Many of the categories were not statistically significant. Of the categories that were statistically significant, the loadings were small. 

## Effect of country of origin factors

As well as the self rated race of the survey participant, the geographical location was also recorded, including the IP address of the country in which the test was conducted. The results were not proportionately distributed amoungst the different countries, with the majority of survey participants coming from the US, Britain, Ireland, and Australia, all native English speaking countries. 
Many of the country factors were not statistically significant.

## Conclusion

something?

## Bibliography

Goldberg, Lewis R. "The development of markers for the Big-Five factor structure." Psychological assessment 4.1 (1992): 26. <http://dx.doi.org/10.1037/1040-3590.4.1.26>

Donnellan, M. B., & Lucas, R. E. (2008). Age Differences in the Big Five Across the Life Span: Evidence from Two National Samples. Psychology and Aging, 23(3), 558–566. http://doi.org/10.1037/a0012897


## Apendix

### Questions asked

#### Personality Questions

E1	I am the life of the party.  
E2	I don't talk a lot.  
E3	I feel comfortable around people.  
E4	I keep in the background.  
E5	I start conversations.  
E6	I have little to say.  
E7	I talk to a lot of different people at parties.  
E8	I don't like to draw attention to myself.  
E9	I don't mind being the center of attention.  
E10	I am quiet around strangers.  
N1	I get stressed out easily.  
N2	I am relaxed most of the time.  
N3	I worry about things.  
N4	I seldom feel blue.  
N5	I am easily disturbed.  
N6	I get upset easily.  
N7	I change my mood a lot.  
N8	I have frequent mood swings.  
N9	I get irritated easily.  
N10	I often feel blue.  
A1	I feel little concern for others.  
A2	I am interested in people.  
A3	I insult people.  
A4	I sympathize with others' feelings.  
A5	I am not interested in other people's problems.  
A6	I have a soft heart.  
A7	I am not really interested in others.  
A8	I take time out for others.  
A9	I feel others' emotions.  
A10	I make people feel at ease.  
C1	I am always prepared.  
C2	I leave my belongings around.  
C3	I pay attention to details.  
C4	I make a mess of things.  
C5	I get chores done right away.  
C6	I often forget to put things back in their proper place.  
C7	I like order.  
C8	I shirk my duties.  
C9	I follow a schedule.  
C10	I am exacting in my work.  
O1	I have a rich vocabulary.  
O2	I have difficulty understanding abstract ideas.  
O3	I have a vivid imagination.  
O4	I am not interested in abstract ideas.  
O5	I have excellent ideas.  
O6	I do not have a good imagination.  
O7	I am quick to understand things.  
O8	I use difficult words.  
O9	I spend time reflecting on things.  
O10	I am full of ideas.  

#### Other Questions

race 

1 = Mixed Race
2 = Arctic (Siberian, Eskimo)
3 = Caucasian (European)
4 = Caucasian (Indian)
5 = Caucasian (Middle East)
6 = Caucasian (North African, Other)
7 = Indigenous Australian
8 = Native American
9 = North East Asian (Mongol, Tibetan, Korean Japanese, etc)
10 = Pacific (Polynesian, Micronesian, etc)
11 = South East Asian (Chinese, Thai, Malay, Filipino, etc)
12 = West African, Bushmen, Ethiopian
13 = Other (0=missed)

engnat	Response to "is English your native language?". 1=yes, 2=no (0=missed)

gender	Chosen from a drop down menu. 1=Male, 2=Female, 3=Other (0=missed)

hand	"What hand do you use to write with?". 1=Right, 2=Left, 3=Both (0=missed)

## R Code

```{r ref.label="exploration", eval=FALSE, echo=TRUE}
```
```{r ref.label="models", eval=FALSE, echo=TRUE}
```
